Texture mapping 3D planar models of indoor environments with noisy camera poses

### 2013-02-04 spie/paper.tex
done with text-based edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{abstract}
  Automated 3D modeling of building interiors is used in applications
  such as virtual reality and environment mapping. Texturing these
  models allows for photo-realistic visualizations of the data
  collected by such modeling systems. While data acquisition times for mobile mapping systems are considerably shorter than for static ones, their recovered camera poses often suffer from inaccuracies, resulting in visible discontinuities when successive images
  are projected onto a surface for texturing. Existing methods to
  stitch images together are often computationally expensive and work
  independently of pose estimates and geometry data.

The approach in this paper minimizes complexity by refining camera poses in 2D, which works well given our high amount of images and minimal 3D pose errors. We also align images to geometry as well as to each other, which produces visually consistent textures even in the presence of inaccurate surface geometry. We also present two different methods for selecting and compositing images into a single texture, depending on the geometry of surfaces being textured. The effectiveness of our method is demonstrated on a number of different indoor environments.

By
  intelligently selecting images for texturing, we develop an
  efficient method to adjust camera poses in 2D, using linear features
  to match images to geometry, and point features to match images
  together. This is followed by two different methods to composite
  images together, based on the geometry of surfaces being
  textured. The effectiveness of our method is demonstrated on a
  number of different indoor environments.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction} % \label{} allows reference to this section
Three-dimensional modeling of indoor environments has a variety of
applications in fields such as architecture, conservation, and
simulation. Applying accurate textures to these models is important to
add context to digital visualizations, and allows for image-based
analysis of environments. In this paper, we perform data acquisition
with a backpack-mounted system carried by an ambulatory human
\cite{liu2010indoor}. This backpack system contains 2 fisheye cameras
as well as laser scanners and inertial measurement units,
used to solve the simultaneous localization and mapping (SLAM)
problem, which produces backpack (and therefore camera) poses over
time as well as a point cloud representing the surrounding environment
\cite{chen2010indoor, liu2010indoor, kua2012loopclosure}. SHOW PICTURE OF BACKPACK. For
environment geometry, we work with models obtained by fitting
low-resolution planar surfaces to these point clouds
\cite{sanchez2012point}.

While a human-carried system provides advantages over more common
wheeled systems in terms of agility and portability, it often leads to
relatively high localization error even after applying sophisticated
localization techniques \cite{liu2010indoor}. Furthermore, our model
geometry usually contains some error as well, and is often tuned to
ignore low-resolution features, such as small walls, furniture, and
other minor non-planar features commonly found in indoor
environments. As a result, methods for texture mapping that rely on
accurate camera poses and projection surfaces generally produce poor
results, while more sophisticated methods to accurately align images
in 3D lead to high runtime on our typically very large datasets. In
contrast, we present a method, which given a high number of images,
searches for a subset of them that can be successfully aligned with 2D
adjustments alone.

The remainder of the paper is organized as follows. Section
\ref{sec:simpleTextureMapping} explains how images are projected onto
our geometry and explains a simple approach towards selecting images
for texturing. Section \ref{sec:existingApproaches} covers existing
approaches to image stitching, and their performance on our
datasets. Section \ref{sec:2dAlignment} contains our approach towards
efficient 2D image alignment, followed by Section
\ref{sec:imageCompositing}, which describes our two methods of
compositing images. Section \ref{sec:resultsAndConclusions} contains
results and conclusions.

@@

The geometry of the texture mapping process for a planar surface is
shown in Figure \ref{fig:projection}.  We are provided with a set of
$M$ images to texture a target plane, where each image has a camera
matrix $P_i$ for $i=1..M$, which translates a 3D point in the world
coordinate system to a 2D point or pixel in image $i$'s coordinates. A
camera matrix $P_i$ is composed of the camera's intrinsic parameters,
containing focal length and image center, as well as extrinsic
parameters which specify the rotation and translation of the camera's
position in 3D world coordinates at the time that image $i$ is
taken. These extrinsic parameters are determined by the backpack
hardware and localization algorithms \cite{chen2010indoor,
  liu2010indoor, kua2012loopclosure} and are substantially noisy.

Because our backpack system takes photos at a rate of 5 Hz, hundreds
of images are available for texturing each surface in our model. Our
goal in designing a texture mapping process is to decide which of
these images should be used, and where their contents should map onto
the final texture, in order to minimize any visual discontinuities or
seams that would suggest that the plane's texture is not composed of a
single continuous image.

@@

\subsection{Tile-based Texture Mapping}
\label{sec:tileBasedMapping}
Ignoring the fact that the camera matrices $P_{1..M}$ are inaccurate,
a simple texturing approach can be performed by discretizing the
target plane into small square tiles, and choosing an image to texture
each tile directly.

We choose to work with rectangular units to ensure that borders
between any two distinct images in the final texture are either
horizontal or vertical. Since most strong environmental features
inside buildings are horizontal or vertical, any visible seams in our
texture will intersect them minimally and be less noticeable.

In order to select an image for texturing a tile $t$, we first gather
a list of candidate images that contain all four of its corners, which
we can quickly check by projecting $t$ into each image using the $P_i$
camera matrices. Furthermore, each candidate image must have been
taken at a time when its camera had a clear line-of-sight to $t$,
which can be calculated using standard ray-polygon intersection tests
between the camera location, $t$, and every other plane
\cite{rayintersection}.

Once we have a list of candidate images for $t$, we define a scoring
function in order to objectively select the best image for texturing
$t$. Since resolution decreases and camera pose errors become more
extreme over distance, we wish to minimize the distance between
cameras and the surfaces they texture. Additionally, we desire images
that are projected perpendicularly onto the plane, maximizing the
resolution and amount of useful texture available in their
projections, as well as minimizing any parallax effects due to
real-world geometry not accurately represented by our digital
model. In other words, we wish to minimize the angle between the
tile's normal vector and the camera axis for images selected for
texturing that tile. These two criteria can be met by maximizing the
function $\frac{1}{d} (-1 \cdot \vec{c}) \cdot \vec{n}$ as shown in
Figure \ref{fig:scoringFunction}. Specifically, $d$ is the distance
between the centers of a camera and a tile, and $\vec{n}$ and
$\vec{c}$ are the directions of the plane's normal and the camera axis
respectively.

As Figure \ref{fig:compareAll}(a) demonstrates, there are many image
boundaries with abrupt discontinuities between tiles, due to
significant misalignment between images, though most of it appears to
be reconcilable using 2D transforms. Given our high number of input
images, a better image selection procedure will further improve
results, as described in Section \ref{sec:imageCompositing}, but a
more significant and reliable improvement can be found through first
improving the alignment of our images.

@@

Feature matching has a number of advantages over direct matching that
make it more suitable for our often non-planar data, and rotational
differences \cite{szeliski2006image}. Feature matching however, works
best when multiple unique visual references exist in the environment
that can be detected in multiple images. In contrast, our indoor
environments have a high prevalence of bare surfaces, as well as
repeating textures, such as similar windows, doors, and wall-mounted
decorations, that cause difficulty in disambiguating features. This
lack of reliable reference points often results in errors when
matching images together.

Additionally, our datasets often contain long chains of images, which
can lead to large overall error when images are matched together. For
example, a slight error in image correspondences in the middle of a
chain could lead to high global error at one or both ends of the
chain.

@@

  \caption{Texture alignment via (a) the graph-based localization
    refinement algorithm, (b) the AutoStitch software package, and (c)
    our method.}
  \label{fig:mosaic3D}
\end{figure}

Past work with our data acquisition backpack integrated image
stitching with an iterative localization algorithm, and performed
homography-based camera pose refinement \cite{liu2010indoor}. When run
on long chains of images, especially where features are sparse, it
produces distorted textures, as seen in Figure \ref{fig:mosaic3D}(a),
due to the problems described above. This approach is also not
closed-form, and its iterative camera adjustment process over our
large datasets led to prohibitively long computation time.

The AutoStitch software package performs homography-based alignment as
well, with additional provisions that attempt to reduce drift and
increase efficiency \cite{panorama2d, autostitch}. Though AutoStitch
also performs well in areas with dense features, it can not handle
areas without features, and has trouble aligning wall sections with
even short segments of bare texture. The example in Figure
\ref{fig:mosaic3D}(b) was generated after manual tuning, and areas
with fewer visual features or repeating texture patterns simply failed
outright with AutoStitch.

More importantly, neither approach

\section{2d Image Alignment}
\label{sec:2dAlignment}

In this section, we describe our method of efficient and robust image
alignment. Though state-of-the-art techniques towards image stitching
work in 3D, we elect to perform image alignments entirely in
2D. Besides the benefits of reduced complexity, we have found 2D
alignment to work well for two main reasons. First, because we have
such a large number of images to choose from, the compositing
approaches in Section \ref{sec:imageCompositing} are able to select a
subset of images that align well, given only 2D alignments. Second,
the nature of our input data is such that localization error chiefly
occurs in two dimensions, in the same plane as the surface being
projected onto. Our data acquisition system contains cameras facing to
the sides of the operator. The operator makes efforts to walk parallel
to walls, and the localization and model-generation algorithms that
provide our input conform the operator's path to be straight and
parallel to detected surfaces \cite{kua2012loopclosure,
  sanchez2012point}. Since the operator also walks in a manner to keep
the backpack balanced, errors in roll and yaw are minimal. Thus, our
highest errors stem from uncertainty in the operator's pitch,
equivalent to rotation around the camera axes, as well as location
along surfaces. These equate to 2D rotation and translation in a
surface's plane.

Our 2D alignment procedure consists of three parts. First, all images
are projected onto the surface and lines within these projections are
detected. These lines are then rotated and shifted to match
geometry-based lines comprising the surface's boundary and
intersection with other surfaces. Second, occlusion checks are
performed to remove invalid parts of each image for the target
surface. These two steps are image-independent and performed in
parallel. Third, we detect SIFT feature matches between pairs of
images and solve a weighted linear least squares problem in 2D to
maximize matches.

\subsection{Geometry-based Alignment}
\label{sec:geometryAlignment}
After computing each image's projection onto the target surface, as
described in Section \ref{sec:simpleTextureMapping}, we detect line
segments in the image projections using Hough transforms, as shown in
Figure \ref{fig:geometryAlignment}(a). Experience and intuition show
that walls in indoor environments often contain linear features that
are either horizontal or vertical, corresponding to doors, windows,
posters, etc. Thus, when texturing walls, we rotate images in 2D such
that dominant lines are made to be horizontal or vertical. This can be
further generalized by instead orienting lines to be parallel or
perpendicular with a surface's boundaries, which makes it applicable
to ceilings and floors, where inner walls and interior features tend
to be parallel or perpendicular to outer walls.

\begin{figure}
  \centering
  \subfloat[][]{\includegraphics[height=3in]{detectedLines.png}}
  \hspace{0.4 cm} \subfloat[][]{\includegraphics[height=3in,
    width=4in]{bothunalignedaligned.png}}
  \caption{(a) Lines detected in image projections via Hough
    transform.  (b) In this example, we project two images at once for
    the sake of demonstration. Red line segments represent where
    surfaces intersect, and green line segments are line segments
    detected in the image. Above are the original projections using
    input camera poses, and below are the projections after rotation
    and translation for alignment.}
  \label{fig:geometryAlignment}
\end{figure}

Furthermore, at this point in time, image occlusions have not been
accounted for. As a result, some image projections contain texture
that should project to an adjacent surface, sometimes with a visible
linear boundary where the two surfaces meet. If this linear boundary
can be detected by Hough transform, the image is rotated and shifted
in 2D such that the visual boundary between two surfaces in an image
projection matches the physical boundary in our digital model. An
example of such an adjustment is in Figure
\ref{fig:geometryAlignment}(b) and \ref{fig:geometryAlignment}(c).

To perform these alignments, we use a RANSAC \cite{fischler1981random}
framework to determine an optimal rotation angle between lines in our
image and lines in our geometry. RANSAC allows us to ignore erroneous
lines in our image, such as the slanted lines corresponding to the
ladder in Figure \ref{fig:geometryAlignment}(a). Once this rotation is
discovered and applied, we gather all pairs of image line segments and
geometry line segments with less than $0.1^\circ$ difference in
orientation and less than 250 mm distance at their furthest points. If
2 or more pairs exist, we select the 2 that contain the longest
noncollinear image-based line segments, and perform the corresponding
fixed translation to match them. If we only have 1 pair, we perform
the minimal shift to match the lines of that pair, and make note of
their orientation, for usage in Section
\ref{sec:robustSIFTFeatureMatching}.

\subsection{Image Occlusion}
\label{sec:imageOcclusion}
Now that images have been aligned to geometry where possible, we
perform a simple recursive occlusion procedure on each image. This is
done by performing the intersection tests in Section
\ref{sec:simpleTextureMapping} in a regularly spaced grid. Where four
corners of a rectangular region are occluded, texture is
removed. Where no corners are occluded, the recursion stops. Where
there is a mixture of both, the rectangular region is subdivided into
four, and the same process is performed on each. By performing Section
\ref{sec:geometryAlignment}'s alignment procedure before occlusion,
texture belonging to other surfaces is accurately removed, which is
necessary for the next section.

\subsection{2D Feature Alignment}
\label{sec:robustSIFTFeatureMatching}
Our next step is to align images by searching for corresponding points
between all pairs of overlapping images. We use feature alignment
rather than pixel or intensity-based alignment due to the differences
in lighting as well as possible occlusion among our images, both of
which feature alignment is less sensitive to \cite{lowe1999object,
  mikolajczyk2005performance, szeliski2006image}.  We use SiftGPU
\cite{siftgpu} for its high performance on both feature detection as
well as pairwise matching. These matches determine $dx$ and $dy$
distances between each pair of features for two image projections,
though these distances may not always be the same for different
features. Since indoor environments often contain repetitive features
such as floor tiles or doors, we need to ensure that SIFT-based
distances are reliable. First, we only perform alignment on parts of
images that overlap given the original noisy poses. Second, we discard
feature matches that correspond to an image distance greater than 200
mm from what the noisy poses estimate. In order to utilize the
remaining feature matches robustly, RANSAC \cite{fischler1981random}
is again used to estimate the optimal $dx_{i,j}$ and $dy_{i,j}$
distances between two images $i$ and $j$. We use a 5 mm threshold for
RANSAC, so that SIFT matches are labeled as outliers if their distance
is not within 5 mm of the sampled average distance.

We now use the $dx_{i,j}$ and $dy_{i,j}$ distances between each pair
of images, as well as our previous geometry alignment results to refine their positions using weighted linear least
squares. The variables we wish to solve for are the $x_i$ and $y_i$
positions of our images, while our equations are the SIFT-based
distances between pairs of images, images fixed to geometry with 0 or
1 degrees of freedom, and the original camera poses. An example setup for
solving our weighted linear least squares problem $\textrm{min}_{\vec{\beta}} ||W^\frac{1}{2}(A \vec{\beta} -
\vec{\gamma})||_2^2 $ with 3 images is shown below.

@@

In this scenario, a SIFT-based distance of $dx_{1,2}$, $dy_{1,2}$ was
calculated between images 1 and 2. This corresponds to the first and
second row of $A$, while the third and fourth row of $A$ represent the
same for images 2 and 3. Row 5 corresponds to a constraint of image 2's location to a line of slope $m_2$, passing through point $gx_2$, $gy_2$. 
Row 6 and 7 do stuff and 8 and 9 ensure that we are not rank deficient in case of whatever.

 Rows 5 and 6 of $A$ correspond to a fixed
location of $tx_1$, $ty_1$ for image 1, while Row 7 corresponds to a
constraint to a line of slope $m$, with current location $tx_2$,
$ty_2$, both possible results of the geometry alignment procedure in
Section \ref{sec:geometryAlignment}. If we do not have enough SIFT
matches, or lack images matched to geometry, our matrix becomes
rank-deficient and our problem cannot be solved. As a result, we add
rows for each image corresponding to the original noisy image
locations, and downweight them heavily, e.g. with a weight of 0.01
versus 1.

@@

\section{Image Compositing}
\label{sec:imageCompositing}
In this section, we revisit the tile-based texturing approach from
Section \ref{sec:simpleTextureMapping}, with an added caching
mechanism to reduce image boundaries. This method works well given all
manner of camera poses and surfaces, but for optimal cases where we
have large sections of usable texture from images, we propose a
superior method that further reduces image boundaries.

@@

As mentioned earlier, our approach is quite efficient. The top wall in
Figure \ref{fig:results}(a) was generated with 7543 $\times$ 776
pixels, and spans a 40-meter long wall. Given 41000 input images of
the entire dataset, a 2.8GHz dual-core consumer-grade laptop takes
approximately a minute to pick 36 candidate images, followed by under
a minute to perform both image alignment and the shortest path
texturing method, though over 75\% of that time is spent calculating
SIFT matches within the SiftGPU framework. While not real-time, the
process is capable of generating quick updates after changes in
various parameters or modifications to input data, and if integrated
directly into a modeling system, could provide live visualization as
data is collected. Our full models consist of our input model file,
textures that we generate, and a mapping of image points to model
vertices. The models shown in Figure \ref{fig:results} are roughly 20
MB in size, and are visualized using the OpenSceneGraph toolkit
\cite{openscenegraph}, which allows for export to many common model
formats, as well as efficient visualization, even in web browsers or
mobile devices.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{abstract}
  Automated 3D modeling of building interiors is used in applications
  such as virtual reality and environment mapping. Texturing these
  models allows for photo-realistic visualizations of the data
  collected by such modeling systems. While data acquisition times for
  mobile mapping systems are considerably shorter than for static
  ones, their recovered camera poses often suffer from inaccuracies,
  resulting in visible discontinuities when successive images are
  projected onto a surface for texturing. Existing methods to stitch
  images together are often computationally expensive and work
  independently of pose estimates and geometry data. We present a
  method for texture mapping that reduces complexity by selecting
  images whose camera poses can be align in 2D. We perform alignment
  of images to geometry as well as images to each other, which
  produces visually consistent textures even in the presence of
  inaccurate surface geometry. We also present two different methods
  for selecting and compositing images into a single texture,
  depending on the geometry of the surface being textured. The
  effectiveness of these methods is demonstrated on a number of different
  indoor environments.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction} % \label{} allows reference to this section
Three-dimensional modeling of indoor environments has a variety of
applications such as training and simulation for disaster management,
virtual heritage conservation, and mapping of hazardous sites. Manual
construction of these digital models can be time consuming, and as
such, automated 3D site modeling has garnered much interest in recent
years.

The first step in automated 3D modeling is the physical scanning of
the environment's geometry. An indoor modeling system must be able to
recover camera poses within an environment while simultaneously
reconstructing the 3D structure of the environment itself
\cite{chen2010indoor, hz, kua2012loopclosure, liu2010indoor}. This is
known as the simultaneous localization and mapping (SLAM) problem, and
is generally solved by taking readings from laser range scanners,
cameras, and inertial measurement units (IMUs) at multiple locations
within the environment.

Mounting such devices on a platform carried by an ambulatory human
provides unique advantages over vehicular-based systems on wheels in
terms of agility and portability, but can also result in larger
localization error \cite{liu2010indoor}. As a result, common methods
for texture mapping generally produce poor results.

In this paper, we present a number of approaches to texture mapping 3D
models of indoor environments made of planar surfaces in the presence
of uncertainty and noise in camera poses. In particular, we consider
data obtained from a human-operated backpack system with a number of
laser range scanners as well as 2 cameras facing left and right, each
equipped with fisheye lenses reaching an approximately 180$^{\circ}$
field of view and taking photos at a rate of 5 Hz. Applying multiple
localization and loop-closure algorithms on the raw data collected by
the onboard sensors \cite{chen2010indoor, kua2012loopclosure,
  liu2010indoor}, the backpack is localized\footnote{In this paper, we
  use the terms localization and pose recovery interchangeably, in
  that they both refer to recovering position and orientation.}  over
its data collection period. This requires recovering the 6 degrees of
freedom for the backpack as well as the cameras rigidly mounted on
it. Once this is complete, the data from the laser range scanners is
used to generate a 3D point cloud of the surrounding environment, from
which a 3D planar model is created \cite{sanchez2012point}. This
model, consisting of 2D polygonal planes in 3D space, along with the
set of images captured by the backpack's cameras and their noisy 3D
poses, can be considered the input to our texture mapping problem.

We perform texture mapping on each planar surface independently and in
parallel. For each surface, we begin by obtaining a set of images that
spans the entire surface with high resolution imagery. We then use our
noisy camera poses to project these selected images onto the
surface. These projections are then refined in 2D, in order to
maximally align them with the surface's geometry, as well as to each
other, allowing us to handle both errors in geometry as well as camera
poses. For surfaces and camera poses at any location or orientation,
we demonstrate a simple tile-based approach for sampling
high-resolution portions of images and compositing them into a
texture. In cases where cameras have consistently perpendicular
viewing angles, we demonstrate a superior approach that leads to more
seamless textures.

The remainder of the paper is organized as follows. Section
\ref{sec:simpleTextureMapping} explains how images are projected onto
geometry and shows a simple texture mapping procedure. Section
\ref{sec:existingApproaches} covers existing approaches to image
stitching, and their performance on our datasets. Section
\ref{sec:2dAlignment} contains our approach towards efficient 2D image
alignment, followed by Section \ref{sec:imageCompositing}, which
describes two methods of selecting and compositing images to create
the final texture. Section \ref{sec:resultsAndConclusions} contains
results and conclusions.

@@

Before describing our texture mapping procedure, we use this section
to explain basic texture mapping and a method for selecting images for
texturing. The geometry of the texture mapping process for a planar
surface is shown in Figure \ref{fig:projection}.  Given a set of $M$
images to texture a target plane, camera matrix $P_i$ for the $ith$
image, translates a 3D point in the world coordinate system to a 2D
point or pixel in image $i$'s coordinates. A camera matrix $P_i$ is
composed of the camera's intrinsic parameters, containing focal length
and image center, as well as extrinsic parameters which specify the
rotation and translation of the camera's position in 3D world
coordinates at the time that image $i$ is taken. These extrinsic
parameters are determined by the backpack hardware and localization
algorithms \cite{chen2010indoor, liu2010indoor, kua2012loopclosure}
and are substantially noisy.

Because the backpack system takes pictures at a rate of 5 Hz, hundreds
of images are available for texturing each surface in the model. Our
objective in designing a texture mapping process is to determine which
of these images should be used, and where their contents should map
onto the final texture, in order to minimize any visual
discontinuities or seams that would suggest that the plane's texture
is not composed of a single continuous image. In the remainder of this
section, we provide an overview of a simple tile-based texture mapping
procedure.

@@

Ignoring the fact that the camera matrices $P_{1..M}$ are inaccurate,
the most straightforward approach is to discretize the target plane
into small square tiles, and choose an image to texture each tile
directly. We choose to work with rectangular units to ensure that
borders between any two distinct images in the final texture are
either horizontal or vertical. Since most strong environmental
features inside buildings are horizontal or vertical, any visible
seams in the texture intersect them minimally and are less noticeable.

In order to select an image for texturing a tile $t$, we first gather
a list of candidate images that contain all four of its corners, which
we can rapidly check by projecting $t$ into each image using the $P_i$
camera matrices. Furthermore, each candidate image must have been
taken at a time when its camera had a clear line-of-sight to $t$,
which can be determined using standard ray-polygon intersection tests
between the camera location, $t$, and every other surface
\cite{rayintersection}.

Once we have a list of candidate images for $t$, we define a scoring
function in order to objectively select the best image for texturing
$t$. Since resolution decreases and camera pose errors become more
pronounced with distance, we wish to minimize the distance between
cameras and the surfaces they texture. Additionally, we desire images
that are projected perpendicularly, rather than obliquely, onto the
plane, maximizing the resolution and amount of useful texture
available in their projections, as well as minimizing any parallax
effects due to real-world geometry not accurately represented by the
digital model. In other words, we wish to minimize the angle between
the tile's normal vector and the camera axis for images selected for
texturing that tile. These two criteria can be met by maximizing the
function $\frac{1}{d} (-1 \cdot \vec{c}) \cdot \vec{n}$ as shown in
Figure \ref{fig:scoringFunction}, where $d$ is the distance between
the centers of a camera and a tile, and $\vec{n}$ and $\vec{c}$ are
the directions of the plane's normal and the camera axis respectively.

As Figure \ref{fig:compareAll}(a) demonstrates, this approach results
in image boundaries with abrupt discontinuities between tiles, due to
significant misalignment between images, though most of it appears to
be reconcilable using 2D transforms. Given the large number of input
images, it is possible to develop a more sophisticated image selection
procedure, as described in Section \ref{sec:imageCompositing},
however, a more significant and reliable improvement can be found
through first improving the alignment of the images as shown in the
next 2 sections.

@@

Feature matching has a number of advantages over direct matching that
make it more suitable for our often non-planar data, and rotational
differences \cite{szeliski2006image}. Feature matching however, works
best when multiple unique visual references exist in the environment
that can be detected in multiple images. In contrast, indoor
environments have a high prevalence of bare surfaces, as well as
repeating textures, such as similar windows, doors, and wall-mounted
decorations, that cause difficulty in disambiguating features. This
lack of reliable reference points often results in errors when
matching images together.

Additionally, our datasets often contain long chains of images,
corresponding to long hallways and corridors as shown in Figure
\ref{fig:mosaic3D}, which leads to error accumulation when image
correspondences are not accurate. For example, when matching a long
chain of images through homography, a pixel in the $nth$ image must be
translated into the first image's coordinates by multiplying by the
$3\times3$ matrix $H_1 H_2 H_3 ... H_n$. Any error in one of these
homography matrices is propagated to all further images, resulting in
drift.

@@

  \caption{Texture alignment via (a) the graph-based localization
    refinement algorithm, (b) the AutoStitch software package, and (c)
    the proposed method.}
  \label{fig:mosaic3D}
\end{figure}

In prior work, we experimented with integrating image stitching with
an iterative localization algorithm \cite{liu2010indoor}. When run on
long chains of images, especially where features are sparse, this
approach produces distorted textures, as seen in Figure
\ref{fig:mosaic3D}(a). Furthermore, this approach is not closed-form,
and its iterative camera adjustment process over large datasets led to
prohibitively long computation time.

The AutoStitch software package performs homography-based alignment as
well, with additional provisions that attempt to reduce drift and
increase efficiency \cite{panorama2d, autostitch}. Though AutoStitch
also performs well in areas with dense features, it can not handle
areas without features, and has trouble aligning wall sections with
even short segments of bare texture. The example in Figure
\ref{fig:mosaic3D}(b) was generated after many rounds of manual
tuning, and areas with fewer visual features or repeating texture
patterns simply failed outright with AutoStitch.

\section{2D Image Alignment}
\label{sec:2dAlignment}
In this section, we describe our method for efficient and robust image
alignment. Rather than attempt to register all of our images in 3D, as
many state-of-the-art techniques for image stitching do, we instead
select a subset of images that can be aligned in 2D. The subset of
images we select for alignment and subsequent texturing corresponds to
the images selected by the tile-based texturing procedure described in
Section \ref{sec:simpleTextureMapping}.

We have found that applying 2D alignments to this set of images works
well for a few reasons. First, the nature of our input data and the
selected images is such that localization error chiefly occurs in two
dimensions, which correspond to the plane of the surface being
projected onto. This is because our backpack operator, during data
acquisition, makes efforts to walk within a few meters of, and
parallel to all walls being scanned. As a result, the translational
error of camera poses is quite minimal in the dimension perpendicular
to the surface being textured. Furthermore, because our set of images
was selected to have minimal distance from our surface, rotational
errors have a lesser effect, except for rotation around the axis
perpendicular to the surface being textured. Therefore, the majority
of both our translation and rotational errors occur in the 2
dimensions of the surface's plane. This of course does not apply for
ceilings and floors; however, the lack of features on most ceilings
and floors means that 3D misalignment does not often have a visible
effect, beyond what can be fixed in 2D.

Our proposed 2D alignment procedure consists of three parts. First, images
are projected onto the surface and lines within these projections are
detected. These lines are then used to transform images such that they match geometry-based lines composing the boundaries of the surface being textured. Second, occlusion checks are performed to remove invalid parts of each image for the target surface. Third, SIFT feature matches are detected between pairs of images, and a weighted linear least squares problem is solved in order to maximize all image and geometry-based alignments. Each step will now be explained in detail.

\subsection{Geometry-based Alignment}
\label{sec:geometryAlignment}

\begin{figure}
  \centering
  \caption{}
  \label{fig:geometryAlignment}
\end{figure}

After computing each image's projection onto the target surface, as
described in Section \ref{sec:simpleTextureMapping}, we obtain a set
of image-based line segments by using Hough transforms to detect lines
in the image projections. We then gather a set of geometry-based
lines, which correspond to the lines comprising the target surface's
border, as well as lines formed where other surfaces intersect the
target surface. An example of these lines is shown for a ceiling
surface in Figure \ref{fig:geometryAlignment}. Given perfect camera
poses and surface geometry, the lines in images corresponding to
corners between surfaces should match up exactly with borders and
intersections of surfaces. By inducing such lines to match, we fit our
camera poses more accurately to our surface, and therefore to
eachother.

To perform geometry-based alignment, we collect pairs of image-based
and geometry-based lines, which are within a distance and orientation
difference threshold of each other. We have found a distance threshold
of 250 mm and an orientation difference threshold of $10^\circ$ to
work well for our datasets. In practice, we rarely have more than 2
such pairs, but if so, we select the 2 noncollinear pairs with the
most similar orientation difference and continue. With 2 noncollinear
pairs, we select the pair with the longer image-based line, and
perform a rotation of the image such that the lines in that pair
become parallel. We then perform a translation such that the lines in
that same pair overlap. This translation has ambiguity in the
dimension along that line, which is resolved by matching the midpoint
of the image-based line in the second pair to its corresponding
geometry-based line. Thus, with 2 or more pairs, we can obtain a fixed
rotation and translation, which we save for usage in Section
\ref{sec:robustSIFTFeatureMatching}. If we only have 1 pair, we
perform a rotation and a minimal translation to match its lines. This
translation's ambiguity however, can not be resolved, but is saved
also to be used in Section
\ref{sec:robustSIFTFeatureMatching}. Finally, in the case where we
have no matches, we can still perform rotations in order to exploit
patterns in indoor environments. For instance, doors, windows,
furniture, etc. all tend to have linear edges that are parallel to the
edges of the surfaces they are on. Similarly, lights, visible interior
walls, etc. which are visible in floor and ceiling images, all tend to
be parallel to exterior walls corresponding to the edges of ceiling
and floor surfaces. Thus, we choose to minimize the angle between
image-based lines and any geometry-based line. We do this using the
RANSAC framework, which successfully ignores outliers
\cite{fischler1981random}.

After these steps are performed, our image projections line up well with our target surface, as shown in Figure \ref{fig:geometryAlignment}. This procedure reconciles both errors in camera poses as well as in geometry, and results in sharp, continuous borders across images, which is crucial when performing occlusion checks.

\subsection{Image Occlusion}
\label{sec:imageOcclusion}
In order to properly texture surfaces such as the one shown in Figure
\ref{fig:occlusion}, it is important to detect when parts of image
projections contain texture belonging to other surfaces and should
therefore not be used. This can be tested by recursively performing
ray-polygon intersection tests between camera locations and other
surfaces in a regularly spaced grid \cite{rayintersection}. Where four
corners of a rectangular region are occluded, texture is
removed. Where no corners are occluded, the recursion stops. Where
there is a mixture of both, the rectangular region is subdivided into
four, and the same process is performed on each. By performing Section
\ref{sec:geometryAlignment}'s alignment procedure before occlusion,
texture belonging to other surfaces is accurately removed, which is
necessary for the next section.

\subsection{2D Feature Alignment}
\label{sec:robustSIFTFeatureMatching}
Our next step is to align images to each other by searching for
corresponding feature points between all pairs of overlapping
images. We use feature alignment rather than pixel or intensity-based
alignment due to the differences in lighting as well as possible
occlusion among our images, both of which feature alignment is less
sensitive to \cite{lowe1999object, mikolajczyk2005performance,
  szeliski2006image}.  We use SiftGPU \cite{siftgpu} for its high
performance on both feature detection as well as pairwise
matching. These matches determine $dx$ and $dy$ distances between each
pair of features for two image projections, though these distances may
not always be the same for different features. Since indoor
environments often contain repetitive features such as floor tiles or
doors, we need to ensure that SIFT-based distances are
reliable. First, we only perform alignment on parts of images that
overlap given the original noisy poses. Second, we discard feature
matches that correspond to an image distance greater than 200 mm from
what the noisy poses estimate. In order to utilize the remaining
feature matches robustly, RANSAC \cite{fischler1981random} is again
used to estimate the optimal $dx_{i,j}$ and $dy_{i,j}$ distances
between two images $i$ and $j$. We use a 5 mm threshold for RANSAC, so
that SIFT matches are labeled as outliers if their distance is not
within 5 mm of the sampled average distance.

We now use the feature-based distances between each pair of images as
well as our previous geometry alignment results to refine all image
positions using a weighted linear least squares solution. The
variables we wish to solve for are the $x_i$ and $y_i$ positions of
images, while equations are the feature-based distances between pairs
of images, images fixed to geometry with 0 or 1 degrees of freedom,
and the original noisy camera poses. The original camera poses are
needed because we sometimes are not able to detect feature matches in
all of our images, or we lack enough geometry alignment results to
generate a single solution. Because we want to minimally use our
original poses, we give them a weighting factor of $0.01$, while all
other equations are weighted at $1$. An example setup for a weighted
linear least squares problem $\textrm{min}_{\vec{\beta}}
||W^\frac{1}{2}(A \vec{\beta} - \vec{\gamma})||_2^2 $ with 3 images is
shown below.

@@

In this scenario, a feature-based distance of $dx_{1,2}$, $dy_{1,2}$
was calculated between images 1 and 2. This corresponds to the first
and second row of $A$, while the third and fourth row of $A$ represent
the same for images 2 and 3. Rows 5 through 7 correspond to results of
the geometry alignment procedure in Section
\ref{sec:geometryAlignment}. Specifically, row 5 corresponds to a
geometry-based constraint of image 2's location to a line of slope
$m_2$, passing through point $gx_2$, $gy_2$, while rows 6 and 7
correspond to a fixed location for image 1 without any degrees of
freedom. Rows 8 and 9 correspond to the original camera pose for image
1, and original poses for images 2 and 3 are omitted for brevity.

@@

\section{Image Compositing}
\label{sec:imageCompositing}
In this section, we revisit the tile-based texturing approach from
Section \ref{sec:simpleTextureMapping}, with an added caching
mechanism to reduce image boundaries. This method works well given all
manner of camera poses and surfaces, but for optimal cases where
images have consistently perpendicular viewing angles, we propose a
superior method that selectively reduces image boundaries.

@@

As mentioned earlier, our approach is quite efficient. The top wall in
Figure \ref{fig:results}(a) was generated with 7543 $\times$ 776
pixels, and spans a 40-meter long wall. Given 41000 input images of
the entire dataset, a 2.8GHz dual-core consumer-grade laptop takes
approximately a minute to pick 36 candidate images, followed by under
a minute to perform both image alignment and the shortest path
texturing method, though over 75\% of that time is spent calculating
SIFT matches within the SiftGPU framework, which could feasibly be
split into a separate preprocessing step. While not real-time, the
process is capable of generating quick updates after changes in
various parameters or modifications to input data, and if integrated
directly into a modeling system, could provide live visualization as
data is collected. Our full models consist of our input model file,
textures that we generate, and a mapping of image points to model
vertices. The models shown in Figure \ref{fig:results} are roughly 20
MB in size, and are visualized using the OpenSceneGraph toolkit
\cite{openscenegraph}, which allows for export to many common model
formats, as well as efficient visualization, even in web browsers or
mobile devices.
